---
title: "Predicting Ductal Invasive Carcinomas via CNN"
author: "Eduardo Coronado"
date: "5/22/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(library(tidyverse))
library(keras)
library(here)
library(curl)
library(png)
suppressMessages(library(filesstrings))
suppressMessages(library(imager))


# Check if data directory exists in current repo, direct user to
# troubleshoot page in case they need to download the data
subdir <- "invasive-carcinoma"
data_dir <- file.path(here(), subdir, "data_idc")
tmp_dir <- file.path(here(), subdir, "tmp")
files_to_keep <-60000 # Define number of patches to keep

if (!file.exists(data_dir)){
  # Download original data zip file - size 1.6 Gb
  download.file("http://andrewjanowczyk.com/wp-static/IDC_regular_ps50_idx5.zip",
                destfile = file.path(here(), subdir, "tmp.zip"))
  
  # Unzip file
  unzip(file.path(here(), subdir, "tmp.zip"), exdir = tmp_dir )
  
  # Take first n 50x50 image patches
  tmp_zero <- list.files(tmp_dir, pattern = "*class0.png", recursive = TRUE,
                           full.names = TRUE)[1:files_to_keep]
  tmp_one <- list.files(tmp_dir, pattern = "*class1.png", recursive = TRUE,
                          full.names = TRUE)[1:files_to_keep]
  
  # Create data directory for 60k patches
  dir.create(data_dir)
  dir.create(file.path(data_dir, "0"))
  dir.create(file.path(data_dir, "1"))
  
  # Moves 60k patches to new dirs
  file.move(tmp_zero[1:60000], destinations =file.path(data_dir, "0") )
  file.move(tmp_one[1:60000], destinations =file.path(data_dir, "1") )
  
  # Clean up
  dir.remove(tmp_dir)
  file.remove(file.path(here(), subdir, "tmp.zip"))
  rm(tmp_zero)
  rm(tmp_one)
}

```

 **_Example and code based on:_**
 
* Baruah, Bikram, and Bikram Baruah. “Predicting Invasive Ductal Carcinoma Using Convolutional Neural Network (CNN) in Keras.” Towards Data Science, Towards Data Science, 3 Jan. 2019, towardsdatascience.com/predicting-invasive-ductal-carcinoma-using-convolutional-neural-network-cnn-in-keras-debb429de9a6. 

* Choosehappy. “Use Case 6: Invasive Ductal Carcinoma (IDC) Segmentation.” Andrew Janowczyk, 5 Jan. 2018, www.andrewjanowczyk.com/use-case-6-invasive-ductal-carcinoma-idc-segmentation/.

```{r}
class_zero <- list.files(data_dir, pattern = "*class0.png", recursive = TRUE, 
                         full.names = TRUE)
class_one <-  list.files(data_dir, pattern = "*class1.png", recursive = TRUE, 
                         full.names = TRUE)


img_width = img_height <- 50


test = readPNG(class_zero, target_size = c(img_width, img_height)) #%>% 
  image_to_array(.) %>% 
  keras::k_expand_dims(., axis = 1) * (1/255)

```